Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2013-06-05 Manuscript/manuscript.tex
Additional methods and results edits
Also changes to R code for highlow analysis -- producing bar chart based
on epidemiological outcome instead of looking at occupancy.

Continuous time stochastic simulations were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).  Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.

@@

The pathogen persisted in 404 of the 1500 simulations, spreading through the entire susceptible host population (i.e. a pandemic) in 147 simulations.  The probability of pathogen persistence increases significantly with pathogen longevity, though patch quality variance has relatively little effect (Figure 1a).  Within cases where the pathogen persists however, the probability of a pandemic increases significantly with both pathogen longevity and patch quality variance (\ref{poutcome}, Figure 1b).

@@

To separate the effects of high and low quality patches in high variance simulations, we selectively replace the ten highest or ten lowest quality patches with patches of unit quality. This reveals that the presence of high quality patches significantly increases infected occupancy, while the presence of low quality patches significantly reduces infected occupancy (Figure 2a).  Similarly, high quality patches decrease susceptible occupancy, while low quality patches increase susceptible occupancy (Figure 2b).  Thus, the increase in the likelihood of a pandemic with increased patch quality variance seems to be attributable to the presence of high quality patches.

\begin{figure}
\label{sens}
\centering
\includegraphics[height=5cm]{highlow(full).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy. The black bar corresponds to a low variance metapopulation.  The \textit{+high} treatment represents the replacement of 10 unit quality patches with 10 high quality patches, while the \textit{+low} treatment represents the replacement of 10 unit quality patches with 10 low quality patches.  Error bars show $\pm$ one standard deviation.}
\end{figure}

@@

\begin{figure}
\label{infections}
\centering
\includegraphics[height=5.5cm]{infevents(full).pdf}
\caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations. Darker shades of blue represent higher point density. The red line represents the loess curve fit to the data.}
\end{figure}

@@

This model framework demonstrates that environmental longevity increases pathogen persistence and spread.  This agrees both with our intuition -- environmental longevity should provide greater opportunity for transmission -- and with models of chronic wasting disease (\cite{Almberg2011},\cite{Sharp2011}).

>>>>

Continuous time stochastic simulations of the above model were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).  Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.

@@

The pathogen persisted in 6954 of the 10000 simulations, spreading through the entire susceptible host population (i.e. a pandemic) in 3734 simulations.  The probability of pathogen persistence increased substantially with pathogen longevity, though patch quality variance had relatively little effect (Figure 1a).  Within cases where the pathogen persisted however, the probability of a pandemic increased with both pathogen longevity and patch quality variance (Figure 1b).

@@

To separate the effects of high and low quality patches, we explore how patch occupancy changes when high and low quality patches are included in the patch quality distribution separately.  In low variance metapopulations (where patch quality ranges from 0.75 to 1.25, corresponding to a variance of 0.02), the proportions of patches occupied by susceptible populations and infected populations are approximately equal, with occupancies of 0.26 and 0.24 respectively.  However, when the patch quality distribution is expanded to include high quality patches (range from 0.75 to 1.75), susceptible occupancy drops to nearly zero, while infected occupancy increases to approximately half of the metapopulation (Figure 2).  Conversely, when the patch quality distribution is expanded to include low quality patches (range from 0.25 to 1.25), susceptible occupancy increases to nearly half, while infected occupancy declines (Figure 2).  When both high and low quality patches are included (range from 0.25 to 1.75, corresponding to a variance of 0.2), their effects approximately cancel, reducing both susceptible and infected occupancy slightly (Figure 2).  Though overall occupancy changes little from low to high variance metapopulations, the proportion of simulations that are pandemics still increases (Figure 1), an effect that can be attributed largely to high quality patches.  However, in high variance metapopulations, the number of disease-free simulations also increases, resulting in relatively little change in mean occupancy.  \emph{Replace these results with outcome-level results for consistency.}

\begin{figure}
\label{sens}
\centering
\includegraphics[height=5cm]{highlow(full).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy. Fix to represent the correct experimental design and results.}
\end{figure}

@@

\begin{figure}
\label{infections}
\centering
\includegraphics[height=7cm]{infevents(full).pdf}
\caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations. Darker shades of blue represent higher point density. The red line represents the loess curve fit to the data.}
\end{figure}

@@

This model framework demonstrates that environmental longevity increases pathogen persistence and spread.  This agrees both with our intuition -- environmental longevity should provide greater opportunity for transmission -- and with models of chronic wasting disease (\cite{Almberg2011},\cite{Sharp2011}).  \emph{Develop further -- both the papers cited, and other studies including a reservoir.  Remember to highlight not just consistency, but also novelty.}
